---
title: "Schwa apocope in Goethe letters"
author: "Michael Cysouw"
date: "`r Sys.Date()`"
output:
  html_document:
    pdf_print: paged
---

```{r}
library(readxl)
library(lme4)
```

# Preparation: data ingestion and cleaning

Read the data from the Excel file. Clean various typos and mark data consistently.

```{r}
data <- read_excel("sources/Schwa annotated data.xls")

data[data=="0.000000"] <- NA
data[data=="?"] <- NA
data[data=="??"] <- NA
data[data=="j "] <- "j"
data[data=="i "] <- "i"
data[data=="t "] <- "t"
data[data=="u "] <- "u"

data[,9:14][data[,9:14] == "0"] <- NA

data$`folgendes Segment`[data$`folgendes Segment`=="I"] <- "i"

data$Sgmentklasse[data$Sgmentklasse=="FSh"] <- "Fsh"
data$Sgmentklasse[data$Sgmentklasse=="PSh"] <- "Psh"
data$Sgmentklasse[data$Sgmentklasse=="v"] <- "V"
data$Sgmentklasse[data$Sgmentklasse=="r"] <- "R"
data$Sgmentklasse[data$Sgmentklasse=="n"] <- "N"
data$Sgmentklasse[data$Sgmentklasse=="l"] <- "L"
data$Sgmentklasse[data$Sgmentklasse=="j"] <- "A"
data$Sgmentklasse[data$Sgmentklasse=="M"] <- "N"

data$Position[data$Position=="v1"] <- "vn"

data$Sgmentklasse[which(data$'folgendes Segment' == "b" & data$Sgmentklasse != "Psh")] <- "Psh"
data$Sgmentklasse[which(data$'folgendes Segment' == "d" & data$Sgmentklasse != "Psh")] <- "Psh"
data$Sgmentklasse[which(data$'folgendes Segment' == "f" & data$Sgmentklasse != "Fsl")] <- "Fsl"
data$Sgmentklasse[which(data$'folgendes Segment' == "h" & data$Sgmentklasse != "Fsl")] <- "Fsl"
data$Sgmentklasse[which(data$'folgendes Segment' == "n" & data$Sgmentklasse != "N")] <- "N"
data$Sgmentklasse[which(data$'folgendes Segment' == "sch" & data$Sgmentklasse != "Fsl")] <- "Fsl"
data$Sgmentklasse[which(data$'folgendes Segment' == "ts" & data$Sgmentklasse != "Psl")] <- "Psl"

data$`Grenze dahinter`[which(data$`Grenze dahinter` > 3)] <- 3
```

# Tables from Section 3

### Table 3: Basic frequencies

```{r}
freq <- table(data$Wort,data$'Schwa?')[c(2,1,3),c(2,1)]
colnames(freq) <- c("with schwa", "without schwa")
# frequencies
freq
# percentages
round(100 * prop.table(freq, 1))
```

## 3.2 Sandhi phonomena

```{r}
three <- data$Sgmentklasse
three[three == "V"] <- "vowel"
three[three %in% c("L","N","R","A")] <- "sonorant"
three[three %in% c("Psh","Psl","Fsh","Fsl")] <- "obstruent" 

t <- table(three, data$`Schwa?`, data$Wort)[3:1,2:1,]
dimnames(t)[[2]] <- c("with schwa", "without schwa")
names(dimnames(t)) <- NULL
```

### Table 4: 'gern(e)'

```{r}
# frequencies
(gerne <- t[,,1])
# standardised residuals
round(chisq.test(gerne)$stdres, digits = 1)
# statistics
chisq.test(gerne)
```

### Table 5: 'heut(e)'

```{r}
# frequencies
(heute <- t[,,2])
# standardised residuals
round(chisq.test(heute)$stdres, digits = 1)
# statistics
chisq.test(heute)
```

## 3.3 Stress clash

```{r}
bet <- table(data$`betonte Silbe folgt`, data$'Schwa?', data$Wort)[2:1,2:1,]
dimnames(bet)[[1]] <- c("unstressed", "stressed")
dimnames(bet)[[2]] <- c("with schwa", "without schwa")
```

### Table 6: significant for 'gern(e)'

```{r}
# frequencies
(gerne <- bet[,,1])
# standardised residuals
round(chisq.test(gerne)$stdres, digits = 1)
# statistics
chisq.test(gerne)
```

### Table 7: not significant for 'heut(e)'

```{r}
# frequencies
(heute <- bet[,,2])
# standardised residuals
round(chisq.test(heute)$stdres, digits = 1)
# statistics
chisq.test(heute)
```

## 3.4 Word Order

```{r}
position <- data$Position
position[position %in% c("v", "vn")] <- "prefield"
position[position == "m1"] <- "beginning of middle field"
position[position == "m2"] <- "middle of middle field"
position[position %in% c("m3", "me","e","n")] <- "end of middle field"

pos <- table(position, data$'Schwa?', data$Wort)[c(4,1,3,2),c(2,1),]
dimnames(pos)[[2]] <- c("with schwa", "without schwa")
names(dimnames(pos)) <- NULL
```

### Table 8: 'gern(e)' (more schwa at end of middle field)

```{r}
# frequencies
(gerne <- pos[,,1])
# standardised residuals
round(chisq.test(gerne)$stdres, digits = 1)
# statistics
chisq.test(gerne)
```

### Table 9: 'heut(e)' (less schwa in prefield)

```{r}
# frequencies
(heute <- pos[,,2])
# standardised residuals
round(chisq.test(heute)$stdres, digits = 1)
# statistics
chisq.test(heute)
```

## 3.5 Prosodic categories and boundaries

```{r}
bound <- table(data$'Grenze dahinter', data$'Schwa?', data$Wort)[c(1,3,2,4),c(2,1),]
dimnames(bound)[[1]] <- 
    c("no boundary", "prefield boundary", "phrase boundary", "utterance boundary")
dimnames(bound)[[2]] <- c("with schwa", "without schwa")
```

### Table 10: 'gern(e)'

```{r}
# frequencies
(gerne <- bound[,,1])
# standardised residuals
round(chisq.test(gerne)$stdres, digits = 1)
# statistics
chisq.test(gerne)
```

### Table 11: 'heut(e)'

```{r}
# frequencies
(heute <- bound[,,2])
# standardised residuals
round(chisq.test(heute)$stdres, digits = 1)
# statistics
chisq.test(heute)
```

## 3.6 Foot structure

```{r}
pref <- table(data$Wort,data$'Schwa?')[c(3,1),c(2,1)]
dimnames(pref)[[2]] <- c("with schwa", "without schwa")
```

### Table 12: 'gern(e)'

```{r}
# frequencies
pref
# standardised residuals
round(chisq.test(pref)$stdres, digits = 1)
# statistics
chisq.test(pref)
```

# Tables from Section 4

Data selection for model

```{r}
dataSel <- data[,c(7,8,10,13,14,15)]
colnames(dataSel) <- c("lexeme", "schwa", "segment", "stress", "position", "boundary")
dataSel <- dataSel[dataSel$lexeme %in% c("gern","heute"),]

# merge all non-vowel segment classes
dataSel$segment[dataSel$segment == "V"] <- "vowel"
dataSel$segment[dataSel$segment != "vowel"] <- "non-vowel"

# merge positions 'beginning' and 'middle' of middlefield
dataSel$position[dataSel$position %in% c("v", "vn")] <- "prefield"
dataSel$position[dataSel$position %in% c("m1", "m2")] <- "beginning middlefield"
dataSel$position[dataSel$position %in% c("m3", "me", "e","n")] <- "end middlefield"

# name boundaries
dataSel$boundary[dataSel$boundary == 0] <- "no boundary"
dataSel$boundary[dataSel$boundary == 1] <- "phrase boundary"
dataSel$boundary[dataSel$boundary == 2] <- "prefield boundary"
dataSel$boundary[dataSel$boundary == 3] <- "utterance boundary"

# turn into factors
dataSel <- data.frame(lapply(dataSel,factor))
levels(dataSel$lexeme) <- c("gerne", "heute")
levels(dataSel$schwa) <- c("no schwa", "with schwa")
levels(dataSel$stress) <- c("stressed", "not stressed")
```

### Table 13: Logistic regression with all two-way interactions.

```{r}
model <- schwa ~ ( lexeme + segment + stress + boundary + position )^2
fit <- step(glm(model, data = dataSel, family = binomial))
knitr::kable(anova (fit, test = "Chi"))
```

It remains unclear what is going on in the interactiond stress:boundary and boundary:position. Some additional calculations that do not occur in the paper. First concerning the interaction stress:boundary, there is a major difference between schwa/no_schwa in interaction with "not stressed" and "utterance boundary". Unclear what this means.

```{r}
# residuals from stress:boundary without schwa
chisq.test(table(dataSel$stress, dataSel$boundary, dataSel$schwa)[,,1])$residuals
# residuals from stress:boundary without schwa
chisq.test(table(dataSel$stress, dataSel$boundary, dataSel$schwa)[,,2])$residuals
```

Second, concerning the interaction position:boundary, there is a major difference between schwa/no_schwa in interaction with "position prefield" and "utterance boundary", but that basically means isolated utterances of adverbs under study. They almost all have a schwa (as might be expected).

```{r}
# residuals from position:boundary without schwa
chisq.test(table(dataSel$position, dataSel$boundary, dataSel$schwa)[,,1])$residuals
# residuals from position:boundary without schwa
chisq.test(table(dataSel$position, dataSel$boundary, dataSel$schwa)[,,2])$residuals
```

## Separate into two regressions

The two words behave really different: therefore we decided to split the data into two regressions and analyze 'gern(e)' and 'heut(e)' separately.

###Table 14/15: 'gern(e)'

There is a preference for no schwa before vowels, before unstressed, and at the start of the middlefield.

```{r}
fitG <- glm(schwa ~ (segment + stress + boundary + position), data = dataSel, family = binomial, subset = lexeme=="gerne")
knitr::kable(anova(fitG, test = "Chi"))
summary(fitG)
```

### Table 16/17: 'heut(e)'

There is a preference for no schwa before vowel, before phrase boundary and before prefield boundare

```{r}
fitH <- glm(schwa ~ segment + stress + boundary + position, data = dataSel, family = binomial, subset = lexeme=="heute")
knitr::kable(anova(fitH, test = "Chi"))
summary(fitH)
```

These are the R-squared values for explained variance, as mentioned at the end of Section 4. They are rather low, but so is the deviance reduction in the anova table. So, these models are not extremely strong in their explanation of the missing schwa. However, for linguistic variation this is not unusual.

```{r}
(gerne <- cor(as.numeric(fitG$model$schwa),predict(fitG, type="link"))^2)
(heute <- cor(as.numeric(fitH$model$schwa),predict(fitH, type="link"))^2)
```

# Mixed models

The following calculations did not end up in the paper, but for people reading this it might of interest. We tried to use a mixed model with the lexeme as a random factor. The results were not very illuminating, though.

```{r}
# preparations
library(lme4)

# on convergence warning: try to get convergence by starting another run from the end result of the previous runs
runAgain <- function(fit) {
	update(fit
		, start = getME( fit, c("theta","fixef") )
		, control = glmerControl( optCtrl = list(maxfun = 2e4) )
		)
}
```

Treating the lexeme as a random factor including all two-way interactions

```{r}
model <- schwa ~ ( segment + stress + boundary + position )^2 + ( 1 | lexeme )
fitRandInter <- glmer(model, data = dataSel, family = binomial)
summary(fitRandInter)
knitr::kable(anova(fitRandInter))
ranef(fitRandInter)
```

Compare different kinds of random factors

```{r}
fitRandBase     <- glmer(schwa ~ segment + stress + boundary + position + ( 1 | lexeme) , data = dataSel, family = binomial)
fitRandSegment  <- glmer(schwa ~ segment + stress + boundary + position + ( 1 + segment | lexeme) , data = dataSel, family = binomial)
fitRandStress   <- glmer(schwa ~ segment + stress + boundary + position + ( 1 + stress | lexeme) , data = dataSel, family = binomial)
fitRandPosition <- glmer(schwa ~ segment + stress + boundary + position + ( 1 + position | lexeme) , data = dataSel, family = binomial)
fitRandBoundary <- glmer(schwa ~ segment + stress + boundary + position + ( 1 + boundary | lexeme) , data = dataSel, family = binomial)

# boundary does not converge, so try to run again
fitRandBoundary <- runAgain(fitRandBoundary)

# not significant: both heute/gerne behave similar
knitr::kable(anova(fitRandBase, fitRandSegment))
# the others are significant: heute/gerne behave different
knitr::kable(anova(fitRandBase, fitRandStress) )
knitr::kable(anova(fitRandBase, fitRandPosition))
knitr::kable(anova(fitRandBase, fitRandBoundary))
```

Direction of random effects is alike to the differences between the two logistic regression models

```{r}
ranef(fitRandSegment)
ranef(fitRandStress)
ranef(fitRandPosition)
ranef(fitRandBoundary)
```

All significant random factors won't converge. Forcing conversion leads to the expected result: segment remains significant, rest is gone.

```{r}
fitRand <- glmer(schwa ~ segment + stress + boundary + position + ( 1 + stress + boundary + position | lexeme) , data = dataSel, family = binomial)
fitRand <- runAgain(fitRand)
summary(fitRand)
```

# Session Info

```{r}
sessionInfo()
```

